Model Finding for Recursive Functions in SMT\thanks{ This research is partially supported by the Inria technological development action ``Contre-exemples Utilisables par Isabelle et Coq'' (CUIC). }

### 2015-05-07 tex/main.tex
more
more refs

@@

\begin{figure}[t]
\normalsize
\begin{enumerate}
%\begin{framed}
\itemx[\ ]
$\conv_0( t\typ{\tau},\, p )$ $=$
\\[-.8\baselineskip]
 \begin{itemize}
   \itemx[] $\mathsf{if}$ $\tau = \Bool$ and $t = \boolop(t_1,\ldots,t_n)$ $\mathsf{then}$
    \begin{itemize}
      \itemx[] $\mathsf{let}$ $( t'_i,\, \chi_i ) = \conv_0( t_i,\, \pol( \boolop,\, i,\, p ) )$ $\mathsf{for}$ $i = 1, \ldots, n$ $\mathsf{in}$%\kern5em$\mid$
      \itemx[] $\mathsf{let}$ $\chi = \chi_1 \wedge \cdots \wedge \chi_n$ $\mathsf{in}$
      \itemx[] $\mathsf{if}$ $p = \ppos$ $\mathsf{then}$
      \begin{itemize}
        \itemx[] \return \bigtuple{$\boolop(t'_1, \ldots, t'_n) ) \wedge \chi,\, \ltrue$}
      \end{itemize}
      \itemx[] $\mathsf{else}$ $\mathsf{if}$ $p = \pneg$ $\mathsf{then}$
      \begin{itemize}
        \itemx[] \return \bigtuple{$\boolop(t'_1, \ldots, t'_n) \vee \neg \chi,\, \ltrue$}
      \end{itemize}
      \itemx[] $\mathsf{else}$
      \begin{itemize}
        \itemx[] \return \bigtuple{$\boolop(t'_1, \ldots, t'_n),\, \chi$}
\\[-.8\baselineskip]
      \end{itemize}
    \end{itemize}
  \itemx[] $\mathsf{else}$ $\mathsf{if}$ $t = \forallf{\con{f}} \vec x.\;\, u$ $\mathsf{then}$
    \begin{itemize}
      \itemx[] $\mathsf{let}$ $( u',\, \chi ) = \conv_0( u,\, p )$ $\mathsf{in}$
      %\itemx[] $\mathsf{assert}( \chi = \ltrue )$;
      \itemx[] \return \bigtuple{$\forall a : \fargtype{\con{f}}.\; u' [ \vecfarg{\con{f}}( a ) / \vec x ],\, \ltrue$}
\\[-.8\baselineskip]
    \end{itemize}
  \itemx[] $\mathsf{else}$ $\mathsf{if}$ $t = \forall \vec x.\;\, u$ $\mathsf{then}$
    \begin{itemize}
      \itemx[] $\mathsf{let}$ $( u',\, \chi ) = \conv_0( u,\, p )$ $\mathsf{in}$
      \itemx[] \return \bigtuple{$\forall \vec x.\; u',\, \forall \vec x.\; \chi$}
\\[-.8\baselineskip]
    \end{itemize}
   \itemx[] $\mathsf{else}$
   \begin{itemize}
     \itemx[] \return \bigtuple{$t,\; \bigwedge\, \{ \exists a : \fargtype{\con{f}}.\; \vecfarg{\con{f}}( a ) \teq \vec t \mid \con{f}( \vec t ) \in \terms( t ),\, \con{f} \in \sfundefs{\Sigma} \}$}
   \end{itemize}
 \end{itemize}
\end{enumerate}
\begin{enumerate}
%\begin{framed}
\itemx[\ ]
$\conv( \varphi )$ $=$
\\[-.8\baselineskip]
 \begin{itemize}
   \itemx[] $\mathsf{let}$ $( \varphi',\, \_ ) = \conv_0( \varphi,\, \ppos )$ $\mathsf{in}$
   %\itemx[] $\mathsf{assert}( D = \emptyset )$;
   \itemx[] \return $\varphi'$
  \end{itemize}
\end{enumerate}
\caption{\,Definition of function $\conv$}
\label{fig:A}
\end{figure}

@@

\begin{lemma}\label{lem:conv}
Let $\varphi$ be a formula not containing function definitions,
and let $\I$ be an $\extendsig{\Sigma}$-interpretation.
Then, $\I$ satisfies $\conv( \varphi )$ if and only if
$\I$ satisfies $L \cup \absconstraints( L )$, where $L$ is a set of ground $\Sigma$-literals that entail $\varphi$.
%There exists a set of ground $\Sigma$-literals $L$ such that $L$ entails $\varphi$, 
%and $\I$ satisfies $L \cup \absconstraints( L )$.
\end{lemma}
\begin{proofsketch}
By definition of $\conv$ and case analysis on the return values of $\conv_0$.
\end{proofsketch}

@@

Second, we show that if $\conv( \varphi )$ is satisfied by a
$\extendsig{\Sigma}$-interpretation $\I$, then $\varphi$ is satisfied by
$\Sigma$-interpretation $\J$. Since $\varphi$ is in definitional form with
respect to the functions defined by $\DDD$, we know $\varphi$ is of the form
$\DDD \wedge \varphi_0$. 
%We argue that $\I$ satisfies a $\Sigma$-formula
%$\varphi'$ that is closed under function expansion with respect to $\DDD$, and
%where $\varphi'$ entails $\varphi_0$.
%To construct $\varphi'$, 
First, we define a sequence of $\Sigma$-literals sets $L_0 \subseteq L_1 \subseteq \ldots$
such that $\I$ satisfies $L_i \cup \absconstraints( L_i )$ for each $i = 0, 1, \ldots$.
Since $\I$ satisfies $\conv( \varphi_0 )$,
by Lemma~\ref{lem:conv}, 
$\I$ satisfies a set of literals $L \cup \absconstraints( L )$ where $L$ is a set of $\Sigma$-literals that entail $\varphi_0$.
Let $L_0 = L$.
For each $i \geq 0$,
let $\psi_i$ be the formula $\bigwedge \{ \conv( \varphi_{\con{f}}[ \vec t / \vec x ] ) \mid \con{f}( \vec t ) \in \terms( L_i ), f \in \sfundefs{\Sigma} \}$,
where the function definition of $\con{f}$ in $\DDD$ is $\forallf{f} \vec x. \varphi_{\con{f}}$.
Since $\I$ satisfies $\conv( \forallf{f} \vec x. \varphi_{\con{f}} )$ and $\absconstraints( L_i )$,
we know that $\I$ also satisfies $\psi_i$.
Thus by Lemma~\ref{lem:conv},
$\I$ satisfies a set of literals $L \cup \absconstraints( L )$ where $L$ is a set of $\Sigma$-literals that entail $\psi_i$.
Let $L_{i+1} = L_0 \cup L$.
Let $L_n$ be the limit of this sequence,
and let $\psi$ be the $\Sigma$-formula $\bigwedge L_n$.
To show $\psi$ is closed under function expansion with respect to $\DDD$,
first note that by construction $\psi$ entails $\psi_n$.
For any function $\con{f}$ and terms $\vec t$, since $\varphi_{\con{f}}[ \vec t / \vec x ]$ does not contain function definitions,
by Corollary~\ref{cor:conv},
$\conv( \varphi_{\con{f}}[ \vec t / \vec x ] )$ entails $\varphi_{\con{f}}[ \vec t / \vec x ]$.
Thus, $\psi$ entails $\{ \varphi_{\con{f}}[ \vec t / \vec x ] \mid \con{f}( \vec t ) \in \terms( \psi ), f \in \sfundefs{\Sigma} \}$,
which by definition implies that $\psi$ is closed under function expansion with respect to $\DDD$.
Furthermore, $\psi$ entails $\varphi_0$ since $L_0 \subseteq L_n$.
Since $\psi$ is a $T$-satisfiable formula that is closed under function expansion with respect to $\DDD$ and $\DDD$ is admissible, 
by definition there exists a $\Sigma$-interpretation $\J$ satisfying $\psi \wedge \DDD$, which entails $\DDD \wedge \varphi_0$,
which is $\varphi$.
\ajr{Sketchy referring to "limit" of sequence, if it can be infinite. TODO: clarify/restrict.}
\qed
\end{proofsketch}

@@

The translation $\conv$ significantly improves the effectiveness of known SMT
techniques for finding counterexamples for formulas involving recursive
function definitions.
For the convenience of end users, it
has been implemented in the latest development version of \cvc %~\cite{ReyEtAl-1-RR-13}
(version~1.5 prerelease).
Function definitions $\forallf{\con{f}} \vec x.\; \varphi$ can be written using the $\definefunreccmd$ command,
from the upcoming SMT-LIB 2.5 standard \cite{smtlib25}.
For example, formula~(\ref{eq:ex-before}) from Example~\ref{ex:translation}
can be specified as
%
\begin{quote}
\small
\begin{verbatim}
(define-fun-rec s ((x Int)) Int (ite (<= x 0) 0 (+ x (s (- x 1)))))
(declare-fun c () Int)
(assert (> (s c) 100))
(check-sat)
\end{verbatim}
\end{quote}
%
When reading this input,
\cvc adds the annotated quantified formula
$$\forallf{\con{s}} x.\;\, \con{s}( x ) \teq \lite\bigl( x \leq 0,\; 0,\; \con{s}( x-1 )\bigr)$$
to its list of assertions,
which after rewriting becomes
$$\forallf{\con{s}} x.\;\, \lite\bigl( x \leq 0,\; \con{s}( x ) \teq 0,\; \con{s}( x ) \teq \con{s}( x-1 ) \bigr)$$
If \cvc's finite model finding mode for recursive functions is enabled (using
the command-line option \texttt{--fmf-fun}), it will replace its list of known
assertions based on the conversion $\conv$ before checking for satisfiability.
%If the functions provided in the input are admissible,
%then a ``satisfiable" response from the solver implies that a model exists for the original formula.
Accordingly, the solver will output the approximation of the interpretation it
used for recursive function definitions. For our example, it will produce the following output:
%\ajr{prefer having the full output like this here?}
%\jb{I like it the way you did it now, but I took the freedom to reformat the output slightly
%(to avoid the small font---I generally try to avoid ad hoc font mixtures---and to show
%the nesting more clearly).}
%it may output the interpretation~(\ref{eq:approx-interp}) from Section~\ref{sec:encoding} for $s$.
%
\begin{quote}
\small
\begin{verbatim}
(model
  (define-fun s (($x1 Int)) Int 
     (ite (= $x1 14) 105 (ite (= $x1 13) 91 (ite (= $x1 12) 78 
        (ite (= $x1 11) 66 (ite (= $x1 10) 55 (ite (= $x1 4) 10 
           (ite (= $x1 9) 45 (ite (= $x1 8) 36 (ite (= $x1 7) 28 
              (ite (= $x1 6) 21 (ite (= $x1 3) 6 (ite (= $x1 5) 15 
                 (ite (= $x1 2) 3 (ite (= $x1 1) 1 0)))))))))))))))
  (define-fun c () Int 14)
)
\end{verbatim}
\end{quote}

The \texttt{--fmf-fun} option instructs \cvc to assume that the
functions provided by the user are admissible.
Admissibility must be checked separately \emph{by the user}---for example, using a
syntactic criterion or a termination prover.
If this criterion is not met, \cvc may answer \emph{sat} on
an unsatisfiable problem. For example, if we add the inconsistent definition
%
\begin{quote}
\small
\begin{verbatim}
(define-fun-rec h ((x Int)) Int (+ (h x) 1))
\end{verbatim}
\end{quote}
%
to the above problem, \cvc will still return ``satisfiable"
if \texttt{--fmf-fun} is specified.

@@

We described the most closely related work, by Ge and de Moura
\cite{GeDeM-CAV-09} and by Reynolds et al.\
\cite{ReyEtAl-1-RR-13,reynolds-et-al-2013}, in the text already.
The finite model finding support in the instantiation-based iProver \cite{xxxx}
is also closery related, given the similarities with SMT.

>>>>

\let\oldcup=\cup
\def\cup{\mathrel{\oldcup}}

@@

\begin{figure}[t]
\normalsize
\begin{enumerate}
%\begin{framed}
\itemx[\ ]
$\conv_0( t\typ{\tau},\, p )$ $=$
\\[-.8\baselineskip]
 \begin{itemize}
   \itemx[] $\mathsf{if}$ $\tau = \Bool$ and $t = \boolop(t_1,\ldots,t_n)$ $\mathsf{then}$
    \begin{itemize}
      \itemx[] $\mathsf{let}$ $( t'_i,\, \chi_i ) = \conv_0( t_i,\, \pol( \boolop,\, i,\, p ) )$ $\mathsf{for}$ $i = 1, \ldots, n$ $\mathsf{in}$%\kern5em$\mid$
      \itemx[] $\mathsf{let}$ $\chi = \chi_1 \wedge \cdots \wedge \chi_n$ $\mathsf{in}$
      \itemx[] $\mathsf{if}$ $p = \ppos$ $\mathsf{then}$
      \begin{itemize}
        \itemx[] \return \bigtuple{$\boolop(t'_1, \ldots, t'_n) ) \wedge \chi,\, \ltrue$}
      \end{itemize}
      \itemx[] $\mathsf{else}$ $\mathsf{if}$ $p = \pneg$ $\mathsf{then}$
      \begin{itemize}
        \itemx[] \return \bigtuple{$\boolop(t'_1, \ldots, t'_n) \vee \neg \chi,\, \ltrue$}
      \end{itemize}
      \itemx[] $\mathsf{else}$
      \begin{itemize}
        \itemx[] \return \bigtuple{$\boolop(t'_1, \ldots, t'_n),\, \chi$}
\\[-.8\baselineskip]
      \end{itemize}
    \end{itemize}
  \itemx[] $\mathsf{else}$ $\mathsf{if}$ $t = \forallf{\con{f}} \vec x.\;\, u$ $\mathsf{then}$
    \begin{itemize}
      \itemx[] $\mathsf{let}$ $( u',\, \chi ) = \conv_0( u,\, p )$ $\mathsf{in}$
      %\itemx[] $\mathsf{assert}( \chi = \ltrue )$;
      \itemx[] \return \bigtuple{$\forall a : \fargtype{\con{f}}.\; u' [ \vecfarg{\con{f}}( a ) / \vec x ],\, \ltrue$}
\\[-.8\baselineskip]
    \end{itemize}
  \itemx[] $\mathsf{else}$ $\mathsf{if}$ $t = \forall \vec x.\;\, u$ $\mathsf{then}$
    \begin{itemize}
      \itemx[] $\mathsf{let}$ $( u',\, \chi ) = \conv_0( u,\, p )$ $\mathsf{in}$
      \itemx[] \return \bigtuple{$\forall \vec x.\; u',\, \forall \vec x.\; \chi$}
\\[-.8\baselineskip]
    \end{itemize}
   \itemx[] $\mathsf{else}$
   \begin{itemize}
     \itemx[] \return \bigtuple{$t,\; \bigwedge\, \{ \exists a : \fargtype{\con{f}}.\; \vecfarg{\con{f}}( a ) \teq \vec t \mid \con{f}( \vec t ) \in \terms( t ),\, \con{f} \in \sfundefs{\Sigma} \}$}
   \end{itemize}
 \end{itemize}
\end{enumerate}
\begin{enumerate}
%\begin{framed}
\itemx[\ ]
$\conv( \varphi )$ $=$
\\[-.8\baselineskip]
 \begin{itemize}
   \itemx[] $\mathsf{let}$ $( \varphi',\, \_ ) = \conv_0( \varphi,\, \ppos )$ $\mathsf{in}$
   %\itemx[] $\mathsf{assert}( D = \emptyset )$;
   \itemx[] \return $\varphi'$
\\[-1.5\baselineskip] %% TYPESETTING
  \end{itemize}
\end{enumerate}
\caption{\,Definition of function $\conv$}
\label{fig:A}
\end{figure}

@@

\begin{lemma}\label{lem:conv}
Let $\varphi$ be a formula not containing function definitions,
and let $\I$ be an $\extendsig{\Sigma}$-interpretation.
Then $\I$ satisfies $\conv( \varphi )$ if and only if
$\I$ satisfies $L \cup \absconstraints( L )$, where $L$ is a set of ground $\Sigma$-literals that entail $\varphi$.
%There exists a set of ground $\Sigma$-literals $L$ such that $L$ entails $\varphi$, 
%and $\I$ satisfies $L \cup \absconstraints( L )$.
\end{lemma}
\begin{proofsketch}
By definition of $\conv$ and case analysis on the return values of $\conv_0$.
\end{proofsketch}

@@

Second, we show that if $\conv( \varphi )$ is satisfied by a
$\extendsig{\Sigma}$-interpretation $\I$, then $\varphi$ is satisfied by
$\Sigma$-interpretation $\J$. Since $\varphi$ is in definitional form with
respect to the functions defined by $\DDD$, we know $\varphi$ is of the form
$\DDD \wedge \varphi_0$. 
%We argue that $\I$ satisfies a $\Sigma$-formula
%$\varphi'$ that is closed under function expansion with respect to $\DDD$, and
%where $\varphi'$ entails $\varphi_0$.
%To construct $\varphi'$, 
First, we define a sequence of $\Sigma$-literals sets $L_0 \subseteq L_1 \subseteq \ldots$
such that $\I$ satisfies $L_i \cup \absconstraints( L_i )$ for each $i = 0, 1, \ldots$.
Since $\I$ satisfies $\conv( \varphi_0 )$,
by Lemma~\ref{lem:conv}, 
$\I$ satisfies a set of literals $L \cup \absconstraints( L )$ where $L$ is a set of $\Sigma$-literals that entail $\varphi_0$.
Let $L_0 = L$.
For each $i \geq 0$,
let $\psi_i$ be the formula $\bigwedge \{ \conv( \varphi_{\con{f}}[ \vec t / \vec x ] ) \mid \con{f}( \vec t ) \in \terms( L_i ), f \in \sfundefs{\Sigma} \}$,
where the function definition of $\con{f}$ in $\DDD$ is $\forallf{f} \vec x. \varphi_{\con{f}}$.
Since $\I$ satisfies $\conv( \forallf{f} \vec x. \varphi_{\con{f}} )$ and $\absconstraints( L_i )$,
we know that $\I$ also satisfies $\psi_i$.
Thus by Lemma~\ref{lem:conv},
$\I$ satisfies a set of literals $L \cup \absconstraints( L )$ where $L$ is a set of $\Sigma$-literals that entail $\psi_i$.
Let $L_{i+1} = L_0 \cup L$.
Let $L_n$ be the limit of this sequence,
and let $\psi$ be the $\Sigma$-formula $\bigwedge L_n$.
To show $\psi$ is closed under function expansion with respect to $\DDD$,
first note that by construction $\psi$ entails $\psi_n$.
For any function $\con{f}$ and terms $\vec t$, since $\varphi_{\con{f}}[ \vec t / \vec x ]$ does not contain function definitions,
by Corollary~\ref{cor:conv},
$\conv( \varphi_{\con{f}}[ \vec t / \vec x ] )$ entails $\varphi_{\con{f}}[ \vec t / \vec x ]$.
Thus, $\psi$ entails $\{ \varphi_{\con{f}}[ \vec t / \vec x ] \mid \con{f}( \vec t ) \in \terms( \psi ), f \in \sfundefs{\Sigma} \}$,
which by definition implies that $\psi$ is closed under function expansion with respect to $\DDD$.
Furthermore, $\psi$ entails $\varphi_0$ since $L_0 \subseteq L_n$.
Since $\psi$ is a $T$-satisfiable formula that is closed under function expansion with respect to $\DDD$ and $\DDD$ is admissible, 
by definition there exists a $\Sigma$-interpretation $\J$ satisfying $\psi \wedge \DDD$, which entails $\DDD \wedge \varphi_0$,
which is $\varphi$.
%\ajr{Sketchy referring to "limit" of sequence, if it can be infinite. TODO: clarify/restrict.}
\qed
\end{proofsketch}

@@

The translation $\conv$ significantly improves the effectiveness of known SMT
techniques for finding counterexamples for formulas involving recursive
function definitions.
For the convenience of end users, it
has been implemented in the latest development version of \cvc %~\cite{ReyEtAl-1-RR-13}
(version~1.5 prerelease).
Function definitions $\forallf{\con{f}} \vec x.\; \varphi$ can be written using the $\definefunreccmd$ command,
from the upcoming SMT-LIB 2.5 standard \cite{smtlib25}.
For example, formula~(\ref{eq:ex-before}) from Example~\ref{ex:translation}
can be specified as
%
\begin{quote}
%\small -- no small, unless you convince me that these mixtures of font sizes
% are aesthetic and make sense --jb
\begin{verbatim}
(define-fun-rec s ((x Int)) Int (ite (<= x 0) 0 (+ x (s (- x 1)))))
(declare-fun c () Int)
(assert (> (s c) 100))
(check-sat)
\end{verbatim}
\end{quote}
%
When reading this input,
\cvc adds the annotated quantified formula
$$\forallf{\con{s}} x.\;\, \con{s}( x ) \teq \lite\bigl( x \leq 0,\; 0,\; \con{s}( x-1 )\bigr)$$
to its list of assertions,
which after rewriting becomes
$$\forallf{\con{s}} x.\;\, \lite\bigl( x \leq 0,\; \con{s}( x ) \teq 0,\; \con{s}( x ) \teq \con{s}( x-1 ) \bigr)$$
If \cvc's finite model finding mode for recursive functions is enabled (using
the command-line option \texttt{--fmf-fun}), it will replace its list of known
assertions based on the conversion $\conv$ before checking for satisfiability.
%If the functions provided in the input are admissible,
%then a ``satisfiable" response from the solver implies that a model exists for the original formula.
Accordingly, the solver will output the approximation of the interpretation it
used for recursive function definitions. For our example, it will produce the following output:
%\ajr{prefer having the full output like this here?}
%\jb{I like it the way you did it now, but I took the freedom to reformat the output slightly
%(to avoid the small font---I generally try to avoid ad hoc font mixtures---and to show
%the nesting more clearly).}
%it may output the interpretation~(\ref{eq:approx-interp}) from Section~\ref{sec:encoding} for $s$.
%
\begin{quote}
%\small
\begin{verbatim}
(model
  (define-fun s (($x1 Int)) Int 
     (ite (= $x1 14) 105 (ite (= $x1 13) 91 (ite (= $x1 12) 78 
        (ite (= $x1 11) 66 (ite (= $x1 10) 55 (ite (= $x1 4) 10 
           (ite (= $x1 9) 45 (ite (= $x1 8) 36 (ite (= $x1 7) 28 
              (ite (= $x1 6) 21 (ite (= $x1 3) 6 (ite (= $x1 5) 15 
                 (ite (= $x1 2) 3 (ite (= $x1 1) 1 0)))))))))))))))
  (define-fun c () Int 14)
)
\end{verbatim}
\end{quote}

The \texttt{--fmf-fun} option instructs \cvc to assume that the
functions provided by the user are admissible.
Admissibility must be checked separately \emph{by the user}---for example, using a
syntactic criterion or a termination prover.
If this criterion is not met, \cvc may answer \emph{sat} on
an unsatisfiable problem. For example, if we add the inconsistent definition
%
\begin{quote}
%\small
\begin{verbatim}
(define-fun-rec h ((x Int)) Int (+ (h x) 1))
\end{verbatim}
\end{quote}
%
to the above problem, \cvc will still return ``satisfiable"
if \texttt{--fmf-fun} is specified.

@@

We described the most closely related work, by Ge and de Moura
\cite{GeDeM-CAV-09} and by Reynolds et al.\
\cite{ReyEtAl-1-RR-13,reynolds-et-al-2013}, in the text already.
The finite model finding support in the instantiation-based iProver
\cite{korovin-2013} is also close, given the similarities with SMT.
